---
title: "CSCI Calculations"
output: html_notebook
---

Source: Instructions for Calculating Bioassessment Indices
Created: October 21st, 2025
Last Updated: December 11th, 2025
Evelyn Malamut, Tahoe Regional Planning Agency
This R Markdown was developed to calculate CSCI scores with yearly bioassessment data
This R script uses R version 4.5.1

Initial Installations and Libraries
```{r}
#Only need to install once
#install.packages("devtools")
#install_github("SCCWRP/BMIMetrics")
#install_github("SCCWRP/CSCI")

library(dplyr)
library(tidyverse)
library(devtools)
library(CSCI)
```

!!USER INPUTS!!
Paths for bug data, station data, and output folders
```{r}
bug_path <- "C:/Users/emalamut/BMI_24_Processing/bug_metrics.csv" #"Record file" tab from "2024 TRPA BMI ABA version analysis"

station_path <- "C:/Users/emalamut/BMI_24_Processing/Indices_Metrics_Consolidated.csv"

#Location to save CSCI outputs
output_path_base <- "C:/Users/emalamut/BMI_24_Processing/"

collection_year <- "2024" #year that you are processing the data for
```

Load, format, and clean bug data
```{r}
#Load and format bug data
bug_metrics <- read_csv(bug_path) %>%
  mutate(
    SampleID = paste(Sample.I.D., collection_year, sep = "_"),
    Distinct = NA
  ) %>%
  select(
    StationCode = Sample.I.D.,
    SampleID,
    FinalID = Taxon,
    LifeStageCode = Stage,
    BAResult = Abundance,
    Distinct
  )

#Load and format station data
stations_CSCI <- read.csv(station_path, stringsAsFactors = F) %>%
  select(
    StationCode, 
    New_Lat, 
    New_Long, 
    SITE_ELEV, 
    ELEV_RANGE, 
    AREA_SQKM, 
    TEMP_00_09, 
    PPT_00_09, 
    SumAve_P, 
    KFCT_AVE, 
    BDH_AVE, 
    P_MEAN
  )

#Clean bug data
bug_clean <- cleanData(bug_metrics)

#List the unique FinalIDs that were flagged as unacceptable values
bug_ID_problems <- bug_clean %>% 
  filter(problemFinalID == TRUE) %>% 
  distinct(FinalID)

bug_ID_problems
```

!!USER INPUT!!
Create list of replacement values for the FinalIDs flagged as unacceptable in the list above (bug_ID_problems). Each unnacceptable value should correspond to an acceptable one. Use the Long output tab in 2024 TRPA BMI ABA version analysis and the SWAMP lookup table for reference. You will need to create your own replacements list. The one below can be used as a reference.
```{r}
replacements <- c(
  "Drunella", 
  "Paraleptophlebia bicornuta", 
  "Lepidostoma", 
  "Physa", 
  "Ephemerella excrucians", 
  "Rhyacophila vemna group", 
  "Perlodidae", 
  "Empididae", 
  "Epeorus longimanus", 
  "Chloroperlidae", 
  "Epeorus grandis", 
  "Tanytarsini", 
  "Turbellaria", 
  "Cricotopus", 
  "Phaenopsectra obediens", 
  "Potthastia longimana group",
  "Lepidostoma",
  "Cricotopus",
  "Leptophlebiidae"
)

#Add a column to the bug_ID_problems dataframe with names of the replacements
bug_ID_problems <- bug_ID_problems %>% 
  mutate(Replacements = replacements)

bug_ID_problems
```


Update bug dataframe with replacements values and check that there are no more unaccepted IDs ( bug_final_Insect_problems should be an empty list)
```{r}
#Replace values in dataframe with replacement values
for (i in 1:length(bug_ID_problems$FinalID)) {
  bug_clean$FinalID <- replace(bug_clean$FinalID, 
                              bug_clean$FinalID == bug_ID_problems[i,1, drop=TRUE], 
                              bug_ID_problems[i,2, drop=TRUE]
                              )
}

#Clean data
bug_final <- cleanData(bug_clean)

#Check if there are still insect ID problems
bug_final_ID_problems <- bug_final %>% 
  filter(problemFinalID == TRUE) %>% 
  distinct(FinalID)

bug_final_ID_problems
```
!!USER CHECK!!
Review list of FinalIDs (bug_insect_problems) flagged as non-insects. If they are all non-insects, then continue.
```{r}
#Select the unique FinalIDs that were flagged as Non-insects
bug_insect_problems <- bug_final %>% 
  filter(LifeStageCode == "X") %>% 
  distinct(FinalID)

bug_insect_problems
```

Generate CSCI report and associated csv files
```{r}
#Calculate CSCI
report <- CSCI(bugs=bug_final, stations=stations_CSCI)

# Save the lists in report as separate CSVs
write.csv(report[["core"]], file = paste0(output_path_base, "CSCI_Report_", collection_year,"_core.csv"))

write.csv(report[["Suppl1_mmi"]], file = paste0(output_path_base, "CSCI_Report_", collection_year,"_Suppl1_mmi.csv"))

write.csv(report[["Suppl2_mmi"]], file = paste0(output_path_base, "CSCI_Report_", collection_year,"_Suppl2_mmi.csv"))

write.csv(report[["Suppl1_grps"]], file = paste0(output_path_base, "CSCI_Report_", collection_year,"_Suppl1_grps.csv"))

write.csv(report[["Suppl1_OE"]], file = paste0(output_path_base, "CSCI_Report_", collection_year,"_Suppl1_OE.csv"))

write.csv(report[["Suppl2_OE"]], file = paste0(output_path_base, "CSCI_Report_", collection_year,"_Suppl2_OE.csv"))
```

